---
title: "_ANTsR_ Bayesian CBF estimation with tissue probabilities"
author: "Brian B. Avants et al."
date: "November 29, 2014"
output:
  pdf_document:
  slidy_presentation:
    self_contained: yes
---

## Basic Goals

- Use bayesian regression to stabilize CBF reconstruction.
- Illustrate the effect of bayesian priors.
- Show the reduction of artifactual effects.
- NOTE: still WIP to incorporate in full pipeline.
- TODO: robust regression params, more locality.
- see [fMRIANTs](https://github.com/stnava/fMRIANTs) for this source file.

## Define study parameters

```{r basics, echo=TRUE,message=FALSE,cache=FALSE}
library(boot)
library(visreg)
library(ANTsR)
library(RKRNS)
library(BMS)
myPriorStrength<-30.0 # controls prior influence
basedir<-"/Users/stnava/data/fMRIANTs/" # FIXME for your study
setwd(basedir)
useDataDrivenMask<-3
denoisingComponents<-1:8
compcorComponents<-0
motionAcc<-1 # motion accuracy - 0 is for testing, 1 or 2 real studies
id<-""; baseSlice<-4 # param for subject 1 
id=2;   baseSlice<-8 # param for subject 2
robustnessvalue<-0.95 # higher rejects more data. 0.9 or less - keep all
seg<-antsImageRead(paste("data3/seg2pcasl",id,".nii.gz",sep=''),3)
tissuelist<-list()
for ( i in c(1:max(seg)) )
  tissuelist[[i]]<-antsImageRead(paste('data3/seg2pcasl',id,'prob',i,'.nii.gz',sep=''),3)
ipttrn<-glob2rx(paste("*PCASL",id,".nii.gz",sep=''))
fns<- paste(basedir,list.files(path=basedir,
  pattern = ipttrn ,recursive=T),sep='/')
if ( all(dim(tissuelist[[1]])==1) 
     | all(dim(seg)==1) |  !file.exists(fns[1]) )
  stop(paste("Check your working directory",basedir))
progbar<-FALSE
hook_plot = knit_hooks$get('plot')
knit_hooks$set(plot = function(x, options) paste('\n', hook_plot(x, options), sep = ''))
```

## Read the image 

```{r reader, echo=TRUE,cache=FALSE}
fn<-fns[1]
if ( ! file.exists(fn) )
  {
  fn<-file.choose()
  }
pcasl<-antsImageRead(fn,4)
```

# A non-expository wrapper function for this project

```{r bcbfun, echo=F,warning=F,message=F,cache=FALSE,eval=FALSE}
bcbf<-bayesianCBF( pcasl, seg, tissuelist,
  myPriorStrength=30.0,
  useDataDrivenMask=3,
  denoisingComponents=1:8,
  robustnessvalue=0.95 ) # higher val up to 1 throws more out
```



## Define output variables

Define output prefix for this subject - usually would contain _unique ID_ concatenated with the _scan date_.

E.g. "110099_20150108".

```{r defout, echo=TRUE,cache=FALSE}
prefix<-paste(tempfile())
figpre<-paste(tempfile())
```


## Get time series average 

```{r getavg, echo=TRUE,warning=F,message=F,cache=FALSE}
avg<-getAverageOfTimeSeries(pcasl)
```

## Show the image 

```{r shower, echo=F,warning=F,message=F,cache=FALSE}
bsl<-baseSlice
myslices<-paste(bsl,bsl+6,2,sep='x')
onm=paste(figpre,'avg.png',sep='')
plotANTsImage(avg,slices=myslices,axis=3,outname=onm)
```
![Axial slices](`r onm`)


## Show Time Slice Before Motion Correction

```{r showmo, echo=FALSE,warning=F,message=F,cache=FALSE}
  boldarr<-as.array(pcasl)
  bold2d<-as.antsImage(t(boldarr[20,20,,]))
  onm2da=paste(figpre,'slices2da.png',sep='')
  plotANTsImage(bold2d,outname=onm2da)
```

![Time Slices Before Motion Correction](`r onm2da`)

## Get masks from data and from T1 map

```{r masker, echo=TRUE,warning=F,message=F,cache=FALSE}
N3BiasFieldCorrection(3,avg,avg,2)
N3BiasFieldCorrection(3,avg,avg,2)
mask<-antsImageClone(seg)
mask[ mask > 0 ]<-1
if ( useDataDrivenMask > 0 )
  {
  mask2<-getMask(avg,mean(avg),Inf,useDataDrivenMask)
  # cleans up mask to agree with data-driven mask2
  mask[mask2==0]<-0 
  seg[mask2==0]<-0
  }
aslmat<-timeseries2matrix(pcasl, mask)
```

## Compute a first pass perfusion map
```{r perfpro, echo=TRUE,warning=F,message=F,cache=FALSE}
  perfpro <- aslPerfusion( pcasl, interpolation="linear", skip=10,
        dorobust=robustnessvalue, useDenoiser=denoisingComponents,  
        moreaccurate=motionAcc, verbose=1, mask=mask, useBayesian=0,
        ncompcor=compcorComponents ) 
  N3BiasFieldCorrection(3,perfpro$m0,perfpro$m0,2)
  pcasl.parameters <- list( sequence="pcasl", m0=perfpro$m0 )
  perfimg<-perfpro$perfusion
```

## Let's look at some interesting voxels

```{r intvox}
intvox<-antsImageRead(paste("data3/PCASL_5_voxel",id,".nii.gz",sep=''),3)
intvoxvec<-intvox[mask==1]
voxlist<-rep(NA,max(intvox))
# 1 - csf, 2 - gm, 3 - wm, 4 - wm & gm , 5 - artifact
voxnm<-c("csf","gm", "wm", "wm/gm","artifact")
for ( i in 1:max(intvox)) {
  voxlist[i]<-which(intvoxvec==as.numeric(i))
  plot(ts(aslmat[,voxlist[i]]),main=voxnm[i])
#  points(mean(aslmat[,voxlist[i]])+perfpro$xideal*10,col='red',type='l')
  }
```

## Let's look at the GM voxel

```{r intvox2}
locdf<-data.frame(voxasl=aslmat[,voxlist[2]], 
                  tc=as.factor(perfpro$xideal), 
                  nuis=perfpro$nuisancevariables)
gmdl<-lm(voxasl~.,data=locdf)
print(summary(gmdl))
visreg(gmdl,'tc')
```

## Bootstrap the TC for the GM voxel

```{r gmboot}
# Bootstrap 95% CI for regression coefficients 
# function to obtain regression weights 
bs <- function(formula, data, indices) {
  d <- data[indices,] # allows boot to select sample 
  fit <- lm(formula, data=d)
  return(coef(fit)) 
} 
# bootstrapping with 1000 replications 
results <- boot(data=locdf, statistic=bs, 
    R=1000, formula=voxasl~.)

# view results
results
plot(results, index=1) # intercept 
plot(results, index=2) # tag control

# get 95% confidence intervals 
boot.ci(results, type="bca", index=2) # tag control 
```

How might we resolve this fairly substantial uncertainty in parameter estimates?

(bayesian model averaging?)

## Let's look at the WM voxel

```{r intvox3}
locdf<-data.frame(voxasl=aslmat[,voxlist[3]], 
                  tc=perfpro$xideal, nuis=perfpro$nuisancevariables)
print(summary(lm(voxasl~.,data=locdf)))
```

## Let's look at the CSF voxel

```{r intvox1}
locdf<-data.frame(voxasl=aslmat[,voxlist[1]], 
                  tc=perfpro$xideal, nuis=perfpro$nuisancevariables)
print(summary(lm(voxasl~.,data=locdf)))
```

## Let's look at a GM/WM voxel

```{r intvox4}
locdf<-data.frame(voxasl=aslmat[,voxlist[4]], 
                  tc=perfpro$xideal, nuis=perfpro$nuisancevariables)
print(summary(lm(voxasl~.,data=locdf)))
```

## Let's look at the artifact voxel

```{r intvox5}
locdf<-data.frame(voxasl=aslmat[,voxlist[5]], 
                  tc=perfpro$xideal, nuis=perfpro$nuisancevariables)
print(summary(lm(voxasl~.,data=locdf)))
nrm<-6
ss<-rep(TRUE,nrow(locdf))
ss[1:(nrm-1)]<-FALSE
ss[(nrow(locdf)-nrm+1):(nrow(locdf))]<-FALSE
print(summary(lm(voxasl~.,data=locdf,subset=ss)))
```


## Get some statistics on which to base priors

```{r baser, echo=TRUE,warning=F,message=F,cache=FALSE}
perfdf<-data.frame( xideal=perfpro$xideal, 
            nuis=perfpro$nuisancevariables)
perfdf<-perfdf[,!is.na(colMeans(perfdf))]
if (  id == "boohoo" ) {
  perfpro$regweights[1:6]<-0
  pp<-length(perfpro$regweights)-6+1
  perfpro$regweights[pp:length(perfpro$regweights)]<-0
}
perfmodel<-lm( aslmat ~.,data=perfdf, weights=perfpro$regweights )
blm<-bigLMStats( perfmodel, includeIntercept=T )
```


## Estimate priors from global map
```{r estimer, echo=TRUE,warning=F,message=F,cache=FALSE}
getpriors<-function( img, seg )
  {
  n<-max(seg)
  p<-rep(0,n)
#  Use the median to be conservative.
  segvec<-( seg[ seg > 0 ] )
  for ( i in 1:n ) p[i]<-median( img[ segvec == as.numeric(i) ] )
  return(p)
  }
bayespriormatfull<-blm$beta
n<-max(seg)*nrow(bayespriormatfull)
bayespriormat<-matrix( rep(0,n), nrow=max(seg) )
for( i in 1:ncol(bayespriormat) )
  bayespriormat[,i]<-getpriors( bayespriormatfull[i,] , seg )
# set 4 to equal 2 - dgm = gm 
bayespriormat[4,]<-bayespriormat[2,]
# set csf to zero perfusion
bayespriormat[1,2]<-0
```

## The prior values for perfusion - per class

```{r printpriors, echo=TRUE,warning=F,message=F,cache=FALSE}
print(bayespriormat[,2])
```

## Extract priors to guide reestimate of perfusion map
```{r extr, echo=TRUE,warning=F,message=F,cache=FALSE}
X<-model.matrix( perfmodel )
localtissuemat<-imageListToMatrix(tissuelist,mask)
priorwt<-diag(ncol(bayespriormat))*myPriorStrength
priorwt[3:ncol(priorwt),3:ncol(priorwt)]<-0
## alternative priors below
# instead use bayespriormatfull - to est cov?
priorwt2<-solve(cov(bayespriormat)+
   diag(ncol(bayespriormat))*.1)*myPriorStrength
```

## Bayesian regression across all tissue models
We use a bayesian model for each tissue i.e. 
$\Pr( \text{tissue}_i | \mathbf{y} )$ where $\mathbf{y}$ is the ASL signal.
Recall the standard approach for estimating perfusion via linear 
regression (forgive brevity):
$$
\mathbf{y} \approx \mathbf{x}^T \mathbf{\beta} + \epsilon,
$$
where $\mathbf{x}$ contains the tag-control designation along 
with other nuisance variables (motion, noise model).
The prior distribution, for each tissue, is
on the parameters of the standard perfusion regression equation which 
we denote as a pair $(\mathbf{\beta}_i,\mathbf{\sigma}^2_i)$ (i.e. mean 
and covariance of the parameters).  The posterior distribution of this 
prior model is proportional to:
$$
\Pr(\mathbf{y}|\mathbf{x},\mathbf{\beta}_i,\mathbf{\sigma}_i^2)
  \Pr(\mathbf{\beta}_i|\mathbf{\sigma}_i^2)\Pr(\mathbf{\sigma}_i^2)
$$
where we will simplify some terms above in the implementation.
We then compute perfusion by integrating over models for all tissues: 
$\sum_i \hat{\beta}_i^p \Pr(\text{tissue}_i)$ where
$\hat{\beta}_i^p$ represents the argmax solution for the 
perfusion given a specific tissue model.  Code below:


```{r bayeservox, echo=TRUE,warning=F,message=F,cache=FALSE}
  for( segval in 1:2 ) {
    localprior<-bayespriormat[segval,]
    blm<-bayesianlm(  X, aslmat[,voxlist[segval]], localprior, priorwt )
    blm2<-bayesianlm(  X, aslmat[,voxlist[segval]], localprior, priorwt*0 )
    blm3<-bayesianlm(  X, aslmat[,voxlist[segval]], localprior, priorwt2 )
    priorwt3<-priorwt2*0 ; diag(priorwt3)<-diag(priorwt2)
    blm4<-bayesianlm(  X, aslmat[,voxlist[segval]], localprior, priorwt3 )
    print(paste(voxnm[segval],blm$beta[1],blm2$beta[1],blm3$beta[1],blm4$beta[1]))
  }
# priorwt<-priorwt2 # consider this ... 
```

Now on the full image.
```{r bayeser, echo=TRUE,warning=F,message=F,cache=FALSE}
bayesianperfusionloc<-localtissuemat*0
bayesianperfusionlocp<-localtissuemat*0
pb <- txtProgressBar(min = 1, max = ncol(aslmat), style = 3)
for ( i in 1:ncol(aslmat) )
  {
  if (progbar)  setTxtProgressBar(pb, i)
  # here is where we get really bayesian 
  # average over all tissue models ...
  localtissuemat[,i]<-abs(localtissuemat[,i])/
    sum(abs(localtissuemat[,i]))
  for ( segval in 1:max(seg) )
    {
    tissueprior<-localtissuemat[segval,i]
    localprior<-bayespriormat[segval,]
    blm<-bayesianlm(  X, aslmat[,i], localprior, priorwt, 
                      regweights=perfpro$regweights )
    locbeta<-blm$beta[1]
    bayesianperfusionloc[segval,i]<-locbeta
    bayesianperfusionlocp[segval,i]<-locbeta*tissueprior
    }
  }
close(pb)
```
This value represents the expected perfusion given the tissue 
content at each voxel.

## Standard vs next-step vs bayesian regularized next-step perfusion map 

```{r bperf, echo=F,warning=F,message=F,cache=FALSE}
bperf<-antsImageClone(mask)
pervec<-colMeans(aslmat[ perfpro$xideal > 0 , ] - aslmat[ perfpro$xideal < 0 , ])
bperf[mask==1]<-pervec
cperfnm=paste(figpre,'perf_crude.png',sep='')
plotANTsImage(bperf,slices=myslices,axis=3,outname=cperfnm)
perfnm=paste(figpre,'perf.png',sep='')
plotANTsImage(perfimg,slices=myslices,axis=3,outname=perfnm)
bperfimg<-makeImage(mask,colSums(bayesianperfusionlocp))
bperfnm=paste(figpre,'bayes_perf.png',sep='')
plotANTsImage(bperfimg,slices=myslices,axis=3,outname=bperfnm)
```



```{r stanvb, echo=F,warning=F,message=F,cache=FALSE}
perfnm=paste(figpre,'perf.png',sep='')
plotANTsImage(perfimg,slices=myslices,axis=3,outname=perfnm)
bperfimg<-makeImage(mask,colSums(bayesianperfusionlocp))
bperfnm=paste(figpre,'bayes_perf.png',sep='')
plotANTsImage(bperfimg,slices=myslices,axis=3,outname=bperfnm)
```
![Axial slices](`r cperfnm`)
![Axial slices](`r perfnm`)
![Axial slices](`r bperfnm`)

## Plot the correlation between bayesian and basic perfusion

```{r plotter, echo=F,warning=F,message=F,cache=FALSE}
plot(  bperf[mask==1],  bperfimg[mask==1], 
  main=paste(cor(bperf[mask==1],  bperfimg[mask==1])) )
```

## M0 image
```{r m0, echo=F,warning=F,message=F,cache=FALSE}
m0nm=paste(figpre,'m0.png',sep='')
plotANTsImage(perfpro$m0,slices=myslices,axis=3,outname=m0nm)
```
![Axial slices](`r m0nm`)

## Quantify CBF
See `aslPerfusion`.
```{r qcbf, echo=TRUE,warning=F,message=F,cache=FALSE}
 # SmoothImage(3,perfpro$m0,3.0,perfpro$m0)
 pcasl.parameters <- list( sequence="pcasl", m0=perfpro$m0 )
 cbf <- quantifyCBF(  bperf, mask,pcasl.parameters )
 bcbf<- quantifyCBF( bperfimg, mask,pcasl.parameters )
```

## Basic and bayesian cbf maps

```{r bbb, echo=F,warning=F,message=F,cache=FALSE}
cbf$kmeancbf[ cbf$kmeancbf < 0 ]<-0
cbfnm=paste(figpre,'cbf.png',sep='')
plotANTsImage(cbf$kmeancbf,slices=myslices,axis=3,outname=cbfnm)
cbf<-cbf$meancbf
bcbf$kmeancbf[ bcbf$kmeancbf < 0 ]<-0
bcbfnm=paste(figpre,'bcbf.png',sep='')
plotANTsImage(bcbf$kmeancbf,slices=myslices,axis=3,outname=bcbfnm)
bcbf<-bcbf$meancbf
```
![Axial slices](`r cbfnm`)
![Axial slices](`r bcbfnm`)


## Statistics: Global Basic CBF
```{r gcb, echo=F,warning=F,message=F,cache=FALSE}
print(paste("Mean",mean(cbf),"sd",sd(cbf),"min",min(cbf),"max",max(cbf)))
for ( i in as.numeric(1:6) ) {
  print(paste("Tissue",i,"Mean",mean(cbf[seg==i]), "sd",sd(cbf[seg==i]),
    "min",min(cbf[seg==i]),"max",max(cbf[seg==i])))
  hist(cbf[seg==i])
  }
```

## Statistics: Global Bayesian CBF
```{r bcb, echo=F,warning=F,message=F,cache=FALSE}
print(paste("Mean",mean(bcbf),"sd",sd(bcbf),"min",min(bcbf),"max",max(bcbf)))
for ( i in as.numeric(1:6) ) {
  print(paste("Tissue",i,"Mean",mean(bcbf[seg==i]), "sd",sd(bcbf[seg==i]),
    "min",min(bcbf[seg==i]),"max",max(bcbf[seg==i])))
  hist(bcbf[seg==i])
  }
```

