---
title: "ANTsR fMRI Questions"
author: "Brian B. Avants et al."
date: "November 24, 2014"
output: 
  pdf_document:
  slidy_presentation:
    self_contained: yes
---

## Basic Goals

- Perform first level statistics for task fMRI with ANTsR
- Map the activation images to the template, assuming `antsCorticalThickness` was run.
- Test on public data, simple task.
- see [fMRIANTs](https://github.com/stnava/fMRIANTs) for this source file.

## Preprocessing for fMRI to perform: Assume co-registration happens later

- Define study parameters
- Read the image 
- Look at it: DVARs
- `antsPreprocessfMRI`:  motion correction, nuisance estimation
- look again
- Define model 
- Do 1st level stats

## Brain mask issues 

- define from BOLD or from act?

## Preprocessing for fMRI to perform: $\beta$s and Co-registration

- Map to subject T1
- ToDo:  Enable imputation etc for dealing with corrupted data

## Define study parameters

```{r, echo=TRUE,message=FALSE,cache=FALSE}
library(ANTsR)
library(RKRNS)
basedir<-"/Users/stnava/data/fMRIANTs/" # FIXME for your study
setwd(basedir)
motionAcc<-0 # motion accuracy - 0 is for testing, 1 or 2 real studies
gtemplate<-antsImageRead("data2/template2mil.nii.gz",3)
t1<-antsImageRead("data2/t1_2mil.nii.gz",3)
ipttrn<-glob2rx(paste("*boldTask.nii.gz",sep=''))
fns<- paste(basedir,list.files(path=basedir,
  pattern = ipttrn ,recursive=T),sep='/')
designfn<-paste(basedir,"data2/designmat.csv",sep='')
if ( all(dim(gtemplate)==1) | ! file.exists(designfn)
     | all(dim(t1)==1) |  !file.exists(fns[1]) ) 
  stop(paste("Check your working directory",basedir))
```

## Define output variables

Define output prefix for this subject - usually would contain _unique ID_ concatenated with the _scan date_.

E.g. "110099_20150108".

```{r, echo=TRUE,cache=FALSE}
prefix<-paste(tempfile())# paste(basedir,"TEMP",sep='') 
figpre<-paste(tempfile())# paste(basedir,"TEMP_fig",sep='')
```


## Quick T1 to template map

```{r, echo=TRUE,message=FALSE,cache=FALSE}
if ( ! exists("templateMaps") )
  {
  templateMaps<-antsRegistration(gtemplate,t1,
    typeofTransform = "SyN",outprefix = prefix)
  antsImageWrite(templateMaps$warpedmovout,
    paste(prefix,'t1warped.nii.gz',sep=''))
  }
```


## Deal with subject specific parameters (AFNI style)

- number of runs?  organize denoising by run etc 
- length of runs
- alternative: estimate polynomial degree of noise? (polort)
- timing files / study design, if it varies per subject
- e.g. look at only correct trials ...


## Read the image 

```{r, echo=TRUE,cache=FALSE}
fn<-fns[1]
if ( ! file.exists(fn) )
  {
  fn<-file.choose()
  }
bold<-antsImageRead(fn,4)
```


## Get time series average 

```{r getavg, echo=TRUE,warning=F,message=F,cache=FALSE}
avg<-getAverageOfTimeSeries(bold)
```

## Show the image 

```{r, echo=F,warning=F,message=F,cache=FALSE}
onm1=paste(figpre,'slices.png',sep='')
plot(avg,slices='18x26x4',axis=3,outname=onm1)
```

## Show the image 

![Axial slices](`r onm1`)


## Get Mask

```{r, echo=TRUE,warning=F,message=F,cache=FALSE}
avg = n3BiasFieldCorrection( avg, 2 )
mask<-getMask(avg,mean(avg),Inf,2)
mat<-timeseries2matrix(bold, mask)
```


## Show the DVARs 

```{r, echo=TRUE,warning=F,message=F,cache=FALSE}
dvars<-computeDVARS(mat)
plot(ts(dvars))
# detect outliers?
```

## Show Time Slice Before Motion Correction

```{r, echo=TRUE,warning=F,message=F,cache=FALSE}
  boldarr<-as.array(bold)
  bold2d<-as.antsImage(t(boldarr[20,20,,]))
  onm2da=paste(figpre,'slices2da.png',sep='')
  plot(bold2d,outname=onm2da)
```

## Show Time Slice Before Motion Correction

![Time Slices Before Motion Correction](`r onm2da`)

## Get TR and ntime volumes 

```{r, echo=TRUE,warning=F,message=F,cache=FALSE}
  tr<-antsGetSpacing(bold)[4]
  nbold<-dim(bold)[4]
```


## Read design file

Contains binary entries indicating event onsets w.r.t. volumes.
```{r, eval=TRUE, cache=FALSE}
  designmat<-read.csv(designfn)
```

## Show design 

```{r, echo=F,warning=F,message=F,eval=TRUE,cache=FALSE}
image(data.matrix(designmat))
```

## Standard preprocessing for fMRI 

What is being done here?

- COMPCOR

- Motion correction

- ... other options

```{r, echo=TRUE,warning=F,message=F,eval=TRUE,cache=FALSE}
  boldpre<-preprocessfMRI( bold, mask, residualizeMatrix=T,
      spatialSmoothingType='gaussian', useMotionCorrectedImage=T,
      motionCorrectionAccuracyLevel = motionAcc,
      spatialSmoothingParameters=0, numberOfCompCorComponents=6 )
```


## Show Time Slice After Motion $+$ Nuisance Residuals

```{r, echo=TRUE,warning=F,message=F,cache=FALSE}
  boldarr<-as.array(boldpre$cleanBoldImage)
  bold2d<-as.antsImage(t(boldarr[20,20,,]))
  onm2db=paste(figpre,'slices2db.png',sep='')
  plot(bold2d,outname=onm2db)
```

## Show Time Slice After Motion $+$ Nuisance Residuals

![Time Slices After Motion Correction](`r onm2db`)

## Look at DVARS in corrected data 

```{r, echo=TRUE,warning=F,message=F,eval=TRUE,cache=FALSE}
plot(ts(boldpre$DVARSpost))
```


## Look at FD in corrected data 

```{r, echo=TRUE,warning=F,message=F,eval=TRUE,cache=FALSE}
plot(ts(boldpre$FD))
```

## Look at COMPCOR or other nuisance variables

Observe the compcor nuisance variables.

```{r, echo=TRUE,warning=F,message=F,eval=TRUE,cache=FALSE}
plot(ts(boldpre$nuisanceVariables[,"compcorr1"]))
```

## Look at COMPCOR or other nuisance variables

Observe the compcor nuisance variables: 2.

```{r, echo=TRUE,warning=F,message=F,eval=TRUE,cache=FALSE}
plot(ts(boldpre$nuisanceVariables[,"compcorr2"]))
```

## Select your nuisance parameters 

```{r, echo=TRUE,warning=F,message=F,eval=TRUE,cache=FALSE}
  print(colnames(boldpre$nuisanceVariables))
  nuis<-cbind(boldpre$nuisanceVariables,
    DVARS=boldpre$DVARS,FD=boldpre$FD)
```

## Smooth and rebuild cleaned BOLD image 

```{r, echo=TRUE,warning=F,message=F,eval=TRUE,cache=FALSE}
  bold<-antsImageClone( boldpre$cleanBoldImage )
  bold = smoothImage( bold, 3.0 )
  mat<-timeseries2matrix( bold, mask  )
```


## Define subject-specific HRF  

Make a block-like function for this design.

```{r, echo=TRUE,warning=F,message=F,eval=TRUE,cache=FALSE}
hrf<-ts( hemodynamicRF(scans=20, onsets=0, durations=12, rt=tr,cc=0.5,a1=8,a2=9))
plot(hrf)
```

## Do convolution with design matrix
```{r, echo=FALSE,eval=TRUE,cache=FALSE}
 hrfdesignmat <- designmat
 doConv = TRUE
 for (i in 1:ncol(hrfdesignmat)) {
   # FIXME - not sure where the function conv lives?
   if ( doConv ) {
     tempx = as.numeric( hrfdesignmat[, i] )
     tempy = as.numeric( hrf )
     hrfdesignmat[, i] <- pracma::conv( tempx, tempy )[1:nrow(hrfdesignmat)]
     }
   }
plot( ts(hrfdesignmat[, 1]) )
```

## Do first level statistics

```{r, echo=TRUE,eval=TRUE,cache=FALSE}
  mdl<-lm( mat ~ A+B+C+D+nuis , data=hrfdesignmat )
  execbetas<-bigLMStats( mdl, 0.0001 )
  print( rownames( execbetas$beta.t ) )
# NOTE - coefficient differences
  betas<-makeImage( mask, 
    execbetas$beta["C",]-execbetas$beta["A",] )
```

## Write some results 

```{r, echo=TRUE,eval=TRUE,cache=FALSE}
  antsImageWrite(betas,paste(prefix,'tbetas.nii.gz',sep=''))
  antsImageWrite(avg,paste(prefix,'avg.nii.gz',sep=''))
  write.csv( nuis, paste(prefix,'nuis.csv',sep=''),row.names=F)
```

## Map to the T1 brain

The `t1brain` : 2mm version here, for speed. 

```{r, echo=TRUE,eval=TRUE,cache=FALSE}
  if (!exists("t1brain")) t1brain<-antsImageClone(t1)
  if (!exists("mytx"))
    {
    mytx<-antsRegistration(fixed=avg, moving=t1brain,
      typeofTransform="SyNBold", 
      outprefix=paste(prefix,"B",sep=''))
    }
```

## Apply mapping to the template through the T1

- note order of transformations given above `antsRegistration` calls

```{r, echo=TRUE,eval=TRUE,cache=FALSE}
  wnm<-templateMaps$fwdtransforms[[1]]
  anm<-templateMaps$fwdtransforms[[2]]
    fulltx<-c(wnm,anm,mytx$invtransforms[1],
      mytx$invtransforms[2])
    wbeta<-antsApplyTransforms(fixed=gtemplate,
      moving=avg, transformlist=fulltx,
      whichtoinvert=c(F,F,T,F) )
    wbetafn<-paste(prefix,'avgw.nii.gz',sep='')
    antsImageWrite(wbeta,wbetafn)
```


## Apply mapping to the template through the T1

- group template is 2mm version

```{r, echo=TRUE,eval=TRUE,cache=FALSE}
    wbeta<-antsApplyTransforms(fixed=gtemplate,
      moving=betas, transformlist=fulltx,
      whichtoinvert=c(F,F,T,F) )
    wbetafn<-paste(prefix,'tbetasw.nii.gz',sep='')
    antsImageWrite(wbeta,wbetafn)
```

## Do group stats

See [ANTsR examples](http://stnava.github.io/ANTsR/)

```{r, echo=TRUE}
# see antsr documentation for how this might be done 
# on the warped beta images
```

## Show on group template 

```{r, echo=T,warning=F,message=F,cache=FALSE}
onm=paste(figpre,'betas.png',sep='')
th<-c(3.0,max(wbeta))
plot(gtemplate,list(wbeta), window.overlay=th,
  color.overlay='red',nslices=20,axis=3 , outname=onm )
```

## Show on group template

![Axial slices](`r onm`)

## $\beta$ histogram

```{r, echo=T}
hist( betas[ mask  == 1  ]   )
```

