---
title: "ANTsR Bayesian CBF Estimation"
author: "Brian B. Avants et al."
date: "November 29, 2014"
output:
  pdf_document:
  slidy_presentation:
    self_contained: yes
---

## Basic Goals

- Use bayesian regression to stabilize CBF reconstruction.
- Illustrate the effect of bayesian priors.
- Show the reduction of artifactual effects.
- NOTE: basic 'beginner' example: still WIP to incorporate in pipeline.
- TODO: add motion correction, robust regression params, locality, use of probabilistic tissue classes to mix models.
- see [fMRIANTs](https://github.com/stnava/fMRIANTs) for this source file.

## Define study parameters

```{r, echo=TRUE,message=FALSE,cache=FALSE}
library(ANTsR)
library(RKRNS)
basedir<-"/Users/stnava/data/fMRIANTs/" # FIXME for your study
setwd(basedir)
motionAcc<-0 # motion accuracy - 0 is for testing, 1 or 2 real studies
gtemplate<-antsImageRead("data2/template2mil.nii.gz",3)
seg<-antsImageRead("data3/seg2pcasl.nii.gz",3)
ipttrn<-glob2rx(paste("*PCASL.nii.gz",sep=''))
fns<- paste(basedir,list.files(path=basedir,
  pattern = ipttrn ,recursive=T),sep='/')
designfn<-paste(basedir,"data2/designmat.csv",sep='')
if ( all(dim(gtemplate)==1) | ! file.exists(designfn)
     | all(dim(seg)==1) |  !file.exists(fns[1]) )
  stop(paste("Check your working directory",basedir))
```

## Define output variables

Define output prefix for this subject - usually would contain _unique ID_ concatenated with the _scan date_.

E.g. "110099_20150108".

```{r, echo=TRUE,cache=FALSE}
myPriorStrength<-50.0 # controls prior influence
prefix<-paste(tempfile())
figpre<-paste(tempfile())
```

Show image.

## Read the image 

```{r, echo=TRUE,cache=FALSE}
fn<-fns[1]
if ( ! file.exists(fn) )
  {
  fn<-file.choose()
  }
pcasl<-antsImageRead(fn,4)
```


## Get time series average 

```{r getavg, echo=TRUE,warning=F,message=F,cache=FALSE}
avg<-getAverageOfTimeSeries(pcasl)
```

## Show the image 

```{r, echo=F,warning=F,message=F,cache=FALSE}
onm=paste(figpre,'avg.png',sep='')
plotANTsImage(avg,slices='6x12x2',axis=3,outname=onm)
```

## Show the image 

![Axial slices](`r onm`)


## Get Mask

```{r, echo=TRUE,warning=F,message=F,cache=FALSE}
N3BiasFieldCorrection(3,avg,avg,2)
N3BiasFieldCorrection(3,avg,avg,2)
mask<-antsImageClone(seg)
mask[ mask > 0 ]<-1
aslmat<-timeseries2matrix(pcasl, mask)
```

Could alternatively use the segmentation mask.

## Compute a first pass perfusion map

```{r, echo=TRUE,warning=F,message=F,cache=FALSE}
tc<-as.factor(rep(c("C","T"),nrow(aslmat)/2))
dv<-computeDVARS(aslmat)
ncompcor<-10
raslmat<-residuals(lm(aslmat~tc+dv))
compcornuis<-compcor(raslmat,ncompcor,variance_extreme = 0.95 )
perfmodel<-lm( aslmat ~ tc + dv + compcornuis )
blm<-bigLMStats( perfmodel, includeIntercept=T )
intvals<-blm$beta["(Intercept)",]
intimg<-makeImage(mask,intvals)
perfvals<-blm$beta["tcT",]
perfimg<-makeImage(mask,perfvals)
dvvals<-blm$beta["dv",]
dvimg<-makeImage(mask,dvvals)
```

## The initial perfusion map 

```{r, echo=F,warning=F,message=F,cache=FALSE}
perfnm=paste(figpre,'perf.png',sep='')
plotANTsImage(perfimg,slices='3x6x1',axis=3,outname=perfnm)
```
![Axial slices](`r perfnm`)

## Estimate priors from global map
```{r, echo=TRUE,warning=F,message=F,cache=FALSE}
getpriors<-function( img, seg )
  {
  n<-max(seg)
  p<-rep(0,n)
  for ( i in 1:n ) p[i]<-median( img[ seg == as.numeric(i) ] )
  return(p)
  }
intpriors<-getpriors( intimg, seg  )
tissuepriors<-getpriors( perfimg, seg )
dvpriors<-getpriors( dvimg, seg )
```
Use the median to be conservative.

## Apply priors to reestimate perfusion map
```{r, echo=TRUE,warning=F,message=F,cache=FALSE}
X<-model.matrix( perfmodel )
segvec<-seg[ mask == 1 ]
bayesianperfusionloc<-rep(0,ncol(aslmat))
for ( i in 1:ncol(aslmat) )
  {
  segval<-segvec[i]
  if (segval==0) segval<-1
  localprior<-c(intpriors[segval], tissuepriors[ segval ], 
                dvpriors[ segval ], rep(0,ncompcor) )
  priorwt<-diag(length(localprior))*myPriorStrength
  # no prior on compcor
  priorwt[4:(4+ncompcor-1),4:(4+ncompcor-1)]<-0 
  blm<-bayesianlm(  X, aslmat[,i], localprior, priorwt )
  bayesianperfusionloc[i]<-blm$beta[1]
  }
```


## The bayesian regularized perfusion map 

```{r, echo=F,warning=F,message=F,cache=FALSE}
bperfimg<-makeImage(mask,bayesianperfusionloc)
bperfnm=paste(figpre,'bayes_perf.png',sep='')
plotANTsImage(bperfimg,slices='3x6x1',axis=3,outname=bperfnm)
```
![Axial slices](`r bperfnm`)

## Plot the correlation between bayesian and basic perfusion

```{r, echo=F,warning=F,message=F,cache=FALSE}
plot(  perfimg[mask==1],  bperfimg[mask==1], 
  main=paste(cor(perfimg[mask==1],  bperfimg[mask==1])) )
```


## Quantify CBF
Note: avg should actually be mean of control images - see `aslPerfusion`.
```{r, echo=TRUE,warning=F,message=F,cache=FALSE}
 pcasl.parameters <- list( sequence="pcasl", m0=avg )
 cbf <- quantifyCBF(  perfimg, mask,pcasl.parameters )
 bcbf<- quantifyCBF( bperfimg, mask,pcasl.parameters )
```

## The basic cbf map 
```{r, echo=F,warning=F,message=F,cache=FALSE}
cbfnm=paste(figpre,'cbf.png',sep='')
plotANTsImage(cbf$kmeancbf,slices='3x6x1',axis=3,outname=cbfnm)
cbf<-cbf$meancbf
```
![Axial slices](`r cbfnm`)


## The bayesian cbf map 

```{r, echo=F,warning=F,message=F,cache=FALSE}
bcbfnm=paste(figpre,'bcbf.png',sep='')
plotANTsImage(bcbf$kmeancbf,slices='3x6x1',axis=3,outname=bcbfnm)
bcbf<-bcbf$meancbf
```
![Axial slices](`r bcbfnm`)


## Statistics: Global Basic CBF
```{r, echo=F,warning=F,message=F,cache=FALSE}
print(paste("Mean",mean(cbf),"sd",sd(cbf),"min",min(cbf),"max",max(cbf)))
for ( i in as.numeric(1:6) )
  print(paste("Tissue",i,"Mean",mean(cbf[seg==i]), "sd",sd(cbf[seg==i]),
    "min",min(cbf[seg==i]),"max",max(cbf[seg==i])))
```

## Statistics: Global Bayesian CBF
```{r, echo=F,warning=F,message=F,cache=FALSE}
print(paste("Mean",mean(bcbf),"sd",sd(bcbf),"min",min(bcbf),"max",max(bcbf)))
for ( i in as.numeric(1:6) )
  print(paste("Tissue",i,"Mean",mean(bcbf[seg==i]), "sd",sd(bcbf[seg==i]),
    "min",min(bcbf[seg==i]),"max",max(bcbf[seg==i])))
```
